String Constraints with Regex-Counting and String-Length Solved Efficiently

### 2023-05-12 2023-1-19/sections/introduction.tex
intro

In modern programming languages---such as JavaScript, Python, Java, and PHP---the string data type plays a crucial role. 
It is well-known that string manipulations are error-prone and could even give rise to security vulnerabilities (e.g. \ cross-site scripting, aka XSS). 
One powerful method for identifying such bugs in programs is \emph{symbolic execution} (possibly in combination with dynamic analysis), which analyses symbolic paths in a program by viewing them as constraints, whose feasibility is checked by constraint solvers. 
%
Symbolic execution of string manipulating programs has motivated the highly active research area of \emph{string constraint solving}, resulting in the development of numerous string solvers in the last decade, including the state-of-the-art string constraint solvers (e.g.
Z3~\cite{Z3}, CVC4/5~\cite{cvc4}, Z3str/2/3/4~\cite{Z3-str,Z3-str2,Z3-str3,BerzishMurphy2021}, Z3str3RE~\cite{BD+23}, 
Z3-Trau~\cite{Z3-trau}, OSTRICH~\cite{CHL+19}, Slent~\cite{WC+18}, among many others.

Regular expressions (abbreviated as regex) and string-length function are widely used in string-manipulating programs. According to the statistics from \cite{CS16,DCSL18,WS18}, regular expressions are used in about 30–40\% Java, JavaScript, and Python software projects. 
Moreover, the statistics from \cite{malware_detection_3_kudzu} show that string-length function occupies 78\% of the occurrences of string operations in 18 Javascript applications. 
%Operations like \verb|length| that take string inputs and return an integer frequently appear in real-world JavaScript applications (78\% of string operations in 18 applications)
As a result, most of the aforementioned string constraint solvers support both regular expressions and string-length function. Moreover, specialized algorithms have been proposed to efficiently solve the class of string constraints with regular expressions and string-length function \cite{LTR+15,BD+23}. 
%a decision procedure based on tableaux calculus was designed in \cite{LTR+15} for the class of string constraints with regular expressions and string-length function. The decision procedure has been implemented in CVC4/5 \cite{LRT+16}.

Counting (aka repetition) is a convenient feature in regular expressions that count the number of matchings of sub-patterns. For instance, $a^{\{2, 4\}}$ specifies that $a$ occurs at least twice and at most four times, and $a^{\{2, \infty\}}$ specifies that $a$ occurs at least twice. 
Note that Kleene star and Kleene plus operator can be seen as special cases of counting operator. For instance, $a^*$ and $a^+$ are equivalent to $a^{\{0,\infty\}}$ and $a^{\{1,\infty\}}$.
Counting is also a frequently used feature of regular expressions. According to the statistics from \cite{CS16}, Kleene star/plus occur in more than 70\% of 1,204 Python projects, while the other forms of counting occur in more than 20\% of them. Therefore, an efficient analysis of string manipulating programs requires efficient solvers for string constraints that contain at least regular expressions with counting and string-length function.

@@

A typical way for string constraint solvers to deal with regex-counting is to unfold regular expressions with counting into those \emph{without} counting by using the concatenation operator. For instance, $a^{\{1, 4\}}$ is unfolded into $a(\varepsilon + a + aa + aaa)$. Since the unfolding incurs an exponential blow-up on the constraint sizes (where the integer constants are encoded in binary), the unfolding penalizes the performance of the solvers, especially when the bounds are large.

@@

\medskip
\noindent 
\emph{Contribution.} In this work, we focus on the class of string constraints with regular membership and string-length function, where the counting operators may occur in regular expressions (called RECL for brevity). We propose an automata-theoretical approach for solving RECL constraints. Our main idea is to represent the counting operators by registers in automata, instead of unfolding them explicitly. Our approach is built upon an automata model called cost-enriched finite automata (CEFA), which was proposed in~\cite{atva2020} to solve string constraints involving the integer data type. Moreover, the string-length function can also be modeled by registers in CEFA. Then the satisfiability of RECL constraints is reduced to the nonemptiness problem of CEFA with respect to a linear integer arithmetic (LIA) formula whose free variables are registers of CEFA.

According to the results from~\cite{atva2020}, an LIA formula can be computed to represent the potential values of registers in CEFA. Thus, the nonemptiness of CEFA w.r.t. LIA formulas can be reduced to the satisfiability of LIA formulas, which is then tackled by off-the-shelf SMT solvers for LIA. 
%
Because the sizes of LIA formulas depend much on the sizes of CEFA, it is beneficial to reduce the sizes of CEFA, in order to achieve a good performance. As a result, we propose various techniques to reduce the state space and the number of transitions for CEFA. Furthermore, since the computation of LIA formulas from CEFA is relatively expensive, we also propose a method of searching for strings that are accepted by CEFA, based on under approximations, before constructing the LIA formulas.

The representation of regex-counting and string-length as registers in CEFA, combined with the aforementioned size-reduction and under-approximation techniques, entails an efficient algorithm for solving RECL constraints. We implement the algorithm on top of OSTRICH, resulting in a new version of OSTRICH, called $\ostrichrecl$ for convenience. 
%
Furthermore, we utilize a collection of benchmark suites, which comprise 48,834 RECL constraints in total, to evaluate the performance of $\ostrichrecl$. The collection of benchmark suites includes a widely-used benchmark suite for string constraints, as well as benchmark suites comprising constraints generated from regular expressions in various open source regex libraries. 
%
%Note that in the benchmark suites, we only consider the regexes that contain explicit counting, that is, regexes that contain sub-expressions of the form $e^{\{n, m\}}$, and ignore those that contain only implicit counting operators, that is, $e^*$ or $e^+$, since the state-of-the-art string solvers suffer mainly from the explicit counting operators. 
%
The experimental results show that $\ostrichrecl$ can solve 48,698 out of 48,834 instances, while the state-of-the-art best solver Z3str3RE on RECL constraints solves only 47,607 instances, with the timeout period set as 60 seconds. In particular, the string constraint in~(\ref{eqn-running}) can be solved by $\ostrichrecl$ within one second. 
Moreover, while $\ostrichrecl$ solves 1,091 more instances than Z3str3RE, it is only slightly slower than it (2.93 versus 1.71 seconds per instance). 
In comparison, CVC5, Z3seq, Z3-Trau, and Z3str3 solve only 44,704, 46,426, 24,121, and 35,866 instances respectively, with an average time 5.67, 3.49, 5.41, and 8.54 seconds per instance. (See Table~\ref{tab:results_all})

At last, we would like to clarify the relationship of this work with~\cite{atva2020}, where CEFA was introduced. This work differs from \cite{atva2020} in the following aspects: In \cite{atva2020}, regex-counting was unfolded explicitly. Moreover, in~\cite{atva2020}, the nonemptiness of CEFA w.r.t. LIA formulas was solved by utilizing the model checker nuXmv~\cite{nuxmv}. In contrast, this work solves it by integrating the techniques of Parikh-image computation, automata-size reduction, and under approximation. The experimental results show that our approach solves 7,092 more instances while spending less time than the solver in~\cite{atva2020}. (See Table~\ref{tab:results_all}, where OSTRICH denotes the solver in \cite{atva2020}.)

\medskip
\noindent
\emph{Related work.} 
We discuss more related work about regexes with counting and real-world regexes in programming languages.  
The determinism of regexes with counting was investigated in~\cite{GGM12,CL15}. Counting-set automata were proposed in~\cite{redos_lenka,HS+23} to match a subclass of regexes with counting quickly. Moreover, a variant of nondeterministic counter automata, called bit vector automata, was proposed recently in \cite{GKM23} to enable fast matching of a more expressive class of regexes with counting.   Nevertheless, the nonemptiness problem of these automata was not considered in ~\cite{redos_lenka,HS+23,GKM23}, and it is unclear whether these automata models can be used for solving string constraints with regex-counting and string-length.
Real-world regexes in programming languages include features beyond classical regular expressions, e.g. the greedy/lazy Kleene star, capturing groups, and back references. Real-world regexes have been addressed in symbolic execution of Javascript programs~\cite{LMK19} and string constraint solving \cite{CF+22}.

@@

\medskip
\noindent
\emph{Organization.} 
The rest of this paper is structured as follows: Section~\ref{sec:overview} gives an overview of the approach in this paper. Section~\ref{sec:pre} introduces the preliminaries. 
Section~\ref{sec:recl} presents the syntax and semantics of RECL. 
Section~\ref{sec:automaton} defines CEFA. Section~\ref{sec:algorithm} introduces the algorithm to solve RECL constraints. Section~\ref{sec:implementation} shows the experiment results. Section~\ref{sec:conclu} concludes this paper.

>>>>

In modern programming languages---such as JavaScript, Python, Java, and PHP---the string data type plays a crucial role. 
It is well-known that string manipulations are error-prone and could even give rise to security vulnerabilities (e.g. \ cross-site scripting, aka XSS). 
One powerful method for identifying such bugs in programs is \emph{symbolic execution} (possibly in combination with dynamic analysis), which analyses symbolic paths in a program by viewing them as constraints, whose feasibility is checked by constraint solvers. 
%
Symbolic execution of string manipulating programs has motivated the highly active research area of \emph{string constraint solving}, resulting in the development of numerous string solvers in the last decade, including the state-of-the-art string constraint solvers, e.g.
Z3seq~\cite{z3seq}, CVC4/5~\cite{cvc4}, Z3str/2/3/4~\cite{Z3-str,Z3-str2,Z3-str3,BerzishMurphy2021}, Z3str3RE~\cite{BD+23}, 
Z3-Trau~\cite{Z3-trau}, OSTRICH~\cite{CHL+19}, Slent~\cite{WC+18}, among many others.

Regular expressions (abbreviated as regex) and string-length function are widely used in string-manipulating programs. According to the statistics from \cite{CS16,DCSL18,WS18}, regular expressions are used in about 30–40\% Java, JavaScript, and Python software projects. 
Moreover, that string-length function occupies 78\% of the occurrences of string operations in 18 Javascript applications, according to the statistics from \cite{malware_detection_3_kudzu}. 
%Operations like \verb|length| that take string inputs and return an integer frequently appear in real-world JavaScript applications (78\% of string operations in 18 applications)
As a result, most of the aforementioned string constraint solvers support both regular expressions and string-length function. Moreover, specialized algorithms have been proposed to efficiently solve the class of string constraints with regular expressions and string-length function \cite{LTR+15,BD+23}. 
%a decision procedure based on tableaux calculus was designed in \cite{LTR+15} for the class of string constraints with regular expressions and string-length function. The decision procedure has been implemented in CVC4/5 \cite{LRT+16}.

Counting (aka repetition) is a convenient feature in regular expressions that count the number of matchings of sub-patterns. For instance, $a^{\{2, 4\}}$ specifies that $a$ occurs at least twice and at most four times, and $a^{\{2, \infty\}}$ specifies that $a$ occurs at least twice. 
Note that Kleene star and Kleene plus operator can be seen as special cases of counting operator. For instance, $a^*$ and $a^+$ are equivalent to $a^{\{0,\infty\}}$ and $a^{\{1,\infty\}}$.
Counting is a frequently used feature of regular expressions. According to the statistics from \cite{CS16}, Kleene star/plus occur in more than 70\% of 1,204 Python projects, while the other forms of counting occur in more than 20\% of them. Therefore, an efficient analysis of string manipulating programs requires efficient solvers for string constraints that contain at least regular expressions with counting and string-length function.

@@

A typical way for string constraint solvers to deal with regex-counting is to unfold regular expressions with counting into those \emph{without} counting by using the concatenation operator. For instance, $a^{\{1, 4\}}$ is unfolded into $a(\varepsilon + a + aa + aaa)$ and $a^{\{2,\infty\}}$ is unfolded into $aaa^{*}$. Since the unfolding incurs an exponential blow-up on the sizes of constraints (assuming that the integer constants in string constraints are encoded in binary), the unfolding penalizes the performance of the solvers, especially when the bounds are large.

@@

\medskip
\noindent 
\emph{Contribution.} In this work, we focus on the class of string constraints with regular membership and string-length function, where the counting operators may occur in regular expressions (called RECL for brevity). We propose an automata-theoretical approach for solving RECL constraints. Our main idea is to represent the counting operators by registers in automata (as much as possible), instead of unfolding them explicitly. Our approach is built upon an automata model called cost-enriched finite automata (CEFA), which was proposed in~\cite{atva2020} to solve string constraints involving the integer data type. Moreover, the string-length function can also be modeled by registers in CEFA. Then the satisfiability of RECL constraints is reduced to the nonemptiness problem of CEFA with respect to a linear integer arithmetic (LIA) formula whose free variables are registers of CEFA.

According to the results from~\cite{atva2020}, an LIA formula can be computed to represent the potential values of registers in CEFA. Thus, the nonemptiness of CEFA w.r.t. LIA formulas can be reduced to the satisfiability of LIA formulas, which is then tackled by off-the-shelf SMT solvers for LIA. 
%
Because the sizes of LIA formulas depend much on the sizes of CEFA, it is beneficial to reduce the sizes of CEFA, in order to achieve a good performance. As a result, we propose various techniques to reduce the state space and the number of transitions of CEFA. Furthermore, since the computation of LIA formulas from CEFA is relatively expensive, we also propose a method of searching for strings that are accepted by CEFA, based on under approximations, before constructing the LIA formulas.

The representation of regex-counting and string-length as registers in CEFA, combined with the aforementioned size-reduction and under-approximation techniques, entails an efficient algorithm for solving RECL constraints. We implement the algorithm on top of OSTRICH, resulting in a new version of OSTRICH, called $\ostrichrecl$ for convenience. 
%
Furthermore, we utilize a collection of benchmark suites, which comprise 48,834 RECL constraints in total, to evaluate the performance of $\ostrichrecl$. The collection of benchmark suites includes a widely-used benchmark suite for string constraints, as well as benchmark suites comprising constraints generated from regular expressions in various open source regex libraries. 
%
%Note that in the benchmark suites, we only consider the regexes that contain explicit counting, that is, regexes that contain sub-expressions of the form $e^{\{n, m\}}$, and ignore those that contain only implicit counting operators, that is, $e^*$ or $e^+$, since the state-of-the-art string solvers suffer mainly from the explicit counting operators. 
%
The experiment results show that $\ostrichrecl$ can solve 48,698 out of 48,834 instances, while the state-of-the-art best solver Z3str3RE on RECL constraints solves only 47,607 instances (where the timeout period is set as 60 seconds). Moreover, while $\ostrichrecl$ solves 1,091 more instances than Z3str3RE, its performance is comparable to Z3str3RE (2.93 versus 1.71 seconds per instance). In comparison, CVC5, Z3seq, Z3-Trau, and Z3str3 solve only 44,704, 46,426, 24,121, and 35,866 instances respectively, with an average time 5.67, 3.49, 5.41, and 8.54 seconds per instance. (See Table~\ref{tab:results_all} for more details.)

At last, we would like to clarify the relationship of this work with~\cite{atva2020}, where CEFA was introduced. This work differs from \cite{atva2020} in the following aspects: In \cite{atva2020}, regex-counting was unfolded explicitly. Moreover, in~\cite{atva2020}, the nonemptiness of CEFA w.r.t. LIA formulas was solved by utilizing the nuXmv model checker~\cite{nuxmv}. In contrast, this work solves it by integrating the techniques of Parikh-image computation, automata-size reduction, and under approximation. The experimental results show that our approach solves 7,092 more instances than the solver in~\cite{atva2020} while spending less time. (See Table~\ref{tab:results_all}, where OSTRICH denotes the solver in \cite{atva2020}.)

\medskip
\noindent
\emph{Related work.} 
We discuss more related work about regexes with counting and real-world regexes in programming languages.  
The determinism of regexes with counting was investigated in~\cite{GGM12,CL15}. Counting-set automata were proposed in~\cite{redos_lenka,HS+23} to match a subclass of regexes with counting quickly. Moreover, a variant of nondeterministic counter automata, called bit vector automata, was proposed recently in \cite{GKM23} to enable fast matching of a more expressive class of regexes with counting.   Nevertheless, the nonemptiness problem of these automata was not considered in ~\cite{redos_lenka,HS+23,GKM23}, and it is unclear whether these automata models can be used for solving string constraints with regex-counting and string-length.
Real-world regexes in programming languages include features beyond classical regular expressions, e.g. the greedy/lazy Kleene star, capturing groups, and back references. Real-world regexes have been addressed in symbolic execution of Javascript programs~\cite{LMK19} and string constraint solving~\cite{CF+22}. Nevertheless, the counting operators are still unfolded explicitly in~\cite{CF+22}.

@@

\smallskip
\noindent
\emph{Organization.} 
The rest of this paper is structured as follows: Section~\ref{sec:overview} gives an overview of the approach in this paper. Section~\ref{sec:pre} introduces the preliminaries. 
Section~\ref{sec:recl} presents the syntax and semantics of RECL. 
Section~\ref{sec:automaton} defines CEFA. Section~\ref{sec:algorithm} introduces the algorithm to solve RECL constraints. Section~\ref{sec:implementation} shows the experiment results. Section~\ref{sec:conclu} concludes this paper.
